# Wrangling 

## Mini Goal
A model having Name, Sex, Mission Year, Occupation, mission title, program selection (or in-orbit), hours_mission

Use a Hierarchical Model with "Selection" as a Group Level?
Account for Non-Independence: Astronauts selected by the same program might share similar training, backgrounds, or selection criteria that influence other variables (e.g., performance, mission roles, outcomes). A hierarchical model can account for this non-independence within groups.

## Occupation cleaning

## Program selection

```{r}
astronauts_df <- astronauts %>%
  select(
    name, mission_title, hours_mission,
    military_civilian, occupation, year_of_mission, in_orbit, sex
  ) %>%
  mutate(
    in_orbit = case_when(
      str_detect(in_orbit, "(?i)^salyut") ~ "Salyut",
      str_detect(in_orbit, "(?i)^apollo") ~ "Apollo",
      str_detect(in_orbit, "(?i)^sts") ~ "STS",
      str_detect(in_orbit, "(?i)^soyuz") ~ "Soyuz",
      str_detect(in_orbit, "(?i)^shenzhou") ~ "Shenzhou",
      str_detect(in_orbit, "(?i)^gemini") ~ "Gemini",
      TRUE ~ in_orbit
    ),
    occupation = str_to_lower(occupation)
  ) %>%
  filter(hours_mission > 0) %>%
  mutate(log_hours_mission = log(hours_mission)) %>%
  na.omit()

in_orbit_cate <- astronauts_df %>% count(in_orbit, sort= TRUE) %>% head(8)

astronauts_df <- astronauts_df %>% filter(
  in_orbit %in% in_orbit_cate$in_orbit
)
  

```

```{r}
astronauts_df
```

```{r}
hiera <- stan_glmer(
  log_hours_mission ~ (1 | in_orbit) + `sex` + occupation, data = astronauts_df,
  family = gaussian,
  chains = 4, iter = 1000*2, seed = 84735, refresh = 0)
# Simulate posterior predictive model of song popularity for each artist
set.seed(84735)
predictions_complete <-  posterior_predict(
  hiera, newdata = astronauts_df)
```


```{r}
pp_check(hiera)
```


```{r}
ggplot(astronauts_df, aes(x = in_orbit, y = hours_mission)) +
  geom_boxplot() +
  labs(title = "Career Length vs. Mission Hours",
       x = "Total Career Hours",
       y = "Mission Duration (hours)")
```

