---
title: "STAT 454 Project Progress Report"
authors: "Khaleesi Chen, Long Truong, Rana Rishmawi"
output: html_document
date: "2024-04-23"
---

## Set up
```{r include = FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  collapse = TRUE, 
  fig.height = 3, 
  fig.width = 5,
  fig.align = 'center')
```


```{r setup, include=FALSE}
library(bayesrules)
library(tidyverse)
library(dplyr)
library(rstanarm)
library(broom.mixed)
library(tidybayes)
library(bayesplot)
astronauts <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-07-14/astronauts.csv')
```

## Research Goals

Analyze the evolution of space missions over time, such as changes in the frequency of missions, duration (hours in mission), and the shift from military to civilian astronauts. This could offer insights into how space exploration has evolved from the Cold War era to the present day. To answer the overarching research question, we would like to concentrate on the mission time total taken based on related variables. 

## Data Background

The dataset, meticulously compiled from various sources including NASA, Roscosmos, and enthusiast-driven websites, offers information of astronautical information. It encapsulates a wide array of data points ranging from the astronauts' full names, gender, birth dates, and nationalities, to their military status and details regarding their selection programs. It is gathered and cleaned by a TidyTuesday contributor named Tom Mock.

### Display information about the dataset
```{r cars}
dim(astronauts)  # Number of rows and columns
names(astronauts)  # First few rows
astronauts %>% head()
```

## Cleaning the Data

```{r pressure, echo=FALSE}

# lowercase the occupation 
astronauts <- astronauts %>%
  mutate(occupation = tolower(occupation))

occupation_counts <- astronauts %>%
  count(occupation, sort = TRUE)

# Convert all in_orbit values to the same title format and remove numbers behind
astronauts <- astronauts %>%
  mutate(in_orbit = gsub("[0-9]", "", in_orbit)) %>%
  mutate(in_orbit = gsub("-", " ", in_orbit))

# Remove extra spaces after "STS" in the in_orbit column
astronauts <- astronauts %>%
  mutate(in_orbit = gsub("STS\\s+", "STS ", in_orbit))


# Remove all non-alphanumeric characters except spaces
astronauts <- astronauts %>%
  mutate(in_orbit = gsub("[^[:alnum:]\\s]", "", in_orbit))

# Remove extra spaces
astronauts <- astronauts %>%
  mutate(in_orbit = gsub("\\s+", " ", in_orbit))

astronauts <- astronauts %>%
  mutate(in_orbit = gsub(".*STS.*", "STS", in_orbit)) %>%
  mutate(in_orbit = gsub(".*Soyuz.*", "Soyuz", in_orbit)) %>%
  mutate(in_orbit = gsub(".*[Aa]pollo.*", "Apollo", in_orbit)) %>%
  mutate(in_orbit = gsub(".*[Gg]emini.*", "Gemini", in_orbit)) %>%
  mutate(in_orbit = gsub(".*Mir.*", "Mir", in_orbit)) %>%
  mutate(in_orbit = gsub(".*Saluyt*", "Salyut", in_orbit)) 


# Filter out rows with 0 hours_mission
astronauts <- astronauts %>%
  filter(hours_mission != 0)

# Log-transform the hours_mission variable
log_transformed_astronauts <- astronauts %>%
  mutate(log_hours_mission = log(hours_mission))  # Adding 1 to avoid log(0)

# factorize desired column vars

log_transformed_astronauts$in_orbit <- as.factor(log_transformed_astronauts$in_orbit)
log_transformed_astronauts$nationality <- as.factor(log_transformed_astronauts$nationality)
log_transformed_astronauts$sex <- as.factor(log_transformed_astronauts$sex)
log_transformed_astronauts$occupation <- as.factor(log_transformed_astronauts$occupation)
log_transformed_astronauts$year_of_selection <- as.factor(log_transformed_astronauts$year_of_selection)
```




## Methodology

- We propose 2 models
- Model 1 is a hierachical model with only a single predictor
- Model 2 is a hierachical model with multiple predictors

## Model  1

### Constructing
```{r}
# hiera_1 <- stan_glmer(
#  log_hours_mission ~ (1 | in_orbit), data = log_transformed_astronauts,
#   family = gaussian,
#   chains = 4, iter = 1000*2, seed = 84735, refresh = 0)

# (complete_pooled, "models/hierachical.rds")

hiera_1 <- readRDS("models/hierachical.rds")
```

### Evaluation
```{r}
astronaut_means <- log_transformed_astronauts %>%
  group_by(in_orbit) %>%
  summarize(count = n(), mission = mean(log_hours_mission))
set.seed(84735)

predictions_complete <-  posterior_predict(
  hiera_1, newdata = astronaut_means)

ppc_intervals(astronaut_means$mission, yrep = predictions_complete,
              prob_outer = 0.80) +
  ggplot2::scale_x_continuous(
    labels = astronaut_means$in_orbit,
    breaks = 1:nrow(astronaut_means)) +
    xaxis_text(angle = 90, hjust = 1)

pp_check(hiera_1)

set.seed(84735)
prediction_summary(model = hiera_1, data =log_transformed_astronauts)
```



### You vs other astronauts
```{r}
set.seed(84735)
predictions_complete_2 <- posterior_predict(
  hiera_1,
  newdata = data.frame(in_orbit = c("You", "STS")))

# Plot the posterior predictive models
mcmc_areas(predictions_complete_2, prob = 0.8) +
  xlab("hours") +
  ggplot2::scale_y_discrete(labels = c("You", "STS"))

```



## Model 2 

### Constructing
```{r}
# hiera <- stan_glmer(
#  log_hours_mission ~ 
#    (1 | in_orbit) + 
#     occupation + 
#    total_number_of_missions +
#    year_of_selection +
#    field21
#  
#  , data = log_transformed_astronauts,
#   family = gaussian,
#   chains = 4, iter = 1000*2, seed = 84735, refresh = 0)

# saveRDS(hiera, "models/new_hierachical.rds")

hiera_2 <- readRDS("models/new_hierachical.rds")
```

### Evaluation
```{r}
pp_check(hiera_2)
```
```{r}
# prediction summary
# hiera_predict_2 <- prediction_summary(model = hiera_2, data=log_transformed_astronauts)
# 
# saveRDS(hiera_predict_2, "hierachical_model_2_predict_cv")
hiera_predict_2
```


