---
title: "STAT 454 Project Progress Report 2"
authors: "Khaleesi Chen, Long Truong, Rana Rishmawi"
output: html_document
date: "2024-04-23"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(bayesrules)
library(tidyverse)
library(dplyr)
library(rstanarm)
library(broom.mixed)
library(tidybayes)
library(bayesplot)
astronauts <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-07-14/astronauts.csv')
```

## Research goals

Analyze the evolution of space missions over time, such as changes in the frequency of missions, duration (hours in mission), and the shift from military to civilian astronauts. This could offer insights into how space exploration has evolved from the Cold War era to the present day. To answer the research question, we would like to concentrate on the mission time total taken based on related variables. Hierarchical modeling is a method that we first think of to tackle the task.

## Data background

The dataset, meticulously compiled from various sources including NASA, Roscosmos, and enthusiast-driven websites, offers information of astronautical information. It encapsulates a wide array of data points ranging from the astronauts' full names, gender, birth dates, and nationalities, to their military status and details regarding their selection programs. It is gathered and cleaned by a TidyTuesday contributor named Tom Mock.

```{r cars}
# Display information about the dataset
dim(astronauts)  # Number of rows and columns
names(astronauts)  # Column names
head(astronauts)  # First few rows
str(astronauts)  # Structure of the dataset
```

## Clean

```{r pressure, echo=FALSE}
astronauts <- astronauts %>%
  mutate(occupation = tolower(occupation))

occupation_counts <- astronauts %>%
  count(occupation, sort = TRUE)

print(occupation_counts)


inorbit_counts <- astronauts %>%
  count(in_orbit, sort = TRUE)

print(inorbit_counts)
unique_inorbit <- unique(astronauts$in_orbit)
print(unique_inorbit)
# Convert all in_orbit values to the same title format and remove numbers behind
astronauts <- astronauts %>%
  mutate(in_orbit = gsub("[0-9]", "", in_orbit)) %>%
  mutate(in_orbit = gsub("-", " ", in_orbit))


# Remove extra spaces after "STS" in the in_orbit column
astronauts <- astronauts %>%
  mutate(in_orbit = gsub("STS\\s+", "STS ", in_orbit))


# Remove all non-alphanumeric characters except spaces
astronauts <- astronauts %>%
  mutate(in_orbit = gsub("[^[:alnum:]\\s]", "", in_orbit))

# Remove extra spaces
astronauts <- astronauts %>%
  mutate(in_orbit = gsub("\\s+", " ", in_orbit))



# Convert any value in the in_orbit column that includes "STS" to "STS"
astronauts <- astronauts %>%
  mutate(in_orbit = gsub(".*STS.*", "STS", in_orbit)) %>%
  mutate(in_orbit = gsub(".*Soyuz.*", "Soyuz", in_orbit)) %>%
  mutate(in_orbit = gsub(".*[Aa]pollo.*", "Apollo", in_orbit)) %>%
  mutate(in_orbit = gsub(".*[Gg]emini.*", "Gemini", in_orbit)) %>%
  mutate(in_orbit = gsub(".*Mir.*", "Mir", in_orbit)) %>%
  mutate(in_orbit = gsub(".*Saluyt*", "Salyut", in_orbit)) 

# Print unique values in the in_orbit column after conversion
unique_inorbit <- unique(astronauts$in_orbit)
print(unique_inorbit)

astronauts %>%
  count(in_orbit, sort = TRUE)
# Define a function to group in_orbit values
group_in_orbit <- function(in_orbit) {
  if (in_orbit %in% c("ISS", "Mir")) {
    return("Group 1")
  } else if (in_orbit %in% c("Salyut", "Skylab")) {
    return("Group 2")
  } else {
    return("Group 3")
  }
}

# Apply the function to create a new column for grouping
astronauts <- astronauts %>%
  mutate(in_orbit_group = sapply(in_orbit, group_in_orbit))

# Check the unique values in the new in_orbit_group column
unique(astronauts$in_orbit_group)
```
```{r}
# Filter out rows with 0 hours_mission
astronauts <- astronauts %>%
  filter(hours_mission != 0)

# Display the first few rows of the filtered dataset
head(astronauts)

# Log-transform the hours_mission variable
log_transformed_astronauts <- astronauts %>%
  mutate(log_hours_mission = log(hours_mission))  # Adding 1 to avoid log(0)

# Display the first few rows of the dataset with the log-transformed variable
head(log_transformed_astronauts)
```
```{r}
model_h <- stan_glm(
  log_hours_mission ~ 1 + in_orbit_group, data = log_transformed_astronauts,
  family = gaussian,
  chains = 4, iter = 5000*2, seed = 84735, refresh = 0)
```

```{r}
pp_check(model_h)
```

```{r}
tidy(model_h, conf.int = TRUE, conf.level = 0.8)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

the hierarchical model from the previous report (report 1) is not good enough so we split the data into 3 groups and we have a hierarchical model for them,the first two groups (group 1 and group 2) last group seems to do well and this is because of the outliers which messes things up. so going forward we will focus on group 3
