---
title: "STAT 454 Project Progress Report 2"
authors: "Khaleesi Chen, Long Truong, Rana Rishmawi"
output: html_document
date: "2024-04-23"
---

```{r setup, include=FALSE}
library(bayesrules)
library(tidyverse)
library(dplyr)
library(rstanarm)
library(broom.mixed)
library(tidybayes)
library(bayesplot)
astronauts <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-07-14/astronauts.csv')
```

## Research Goals

Analyze the evolution of space missions over time, such as changes in the frequency of missions, duration (hours in mission), and the shift from military to civilian astronauts. This could offer insights into how space exploration has evolved from the Cold War era to the present day. To answer the research question, we would like to concentrate on the mission time total taken based on related variables. Hierarchical modeling is a method that we first think of to tackle the task.

## Data Background

The dataset, meticulously compiled from various sources including NASA, Roscosmos, and enthusiast-driven websites, offers information of astronautical information. It encapsulates a wide array of data points ranging from the astronauts' full names, gender, birth dates, and nationalities, to their military status and details regarding their selection programs. It is gathered and cleaned by a TidyTuesday contributor named Tom Mock.

### Display information about the dataset
```{r cars}
dim(astronauts)  # Number of rows and columns
names(astronauts)  # First few rows
```

## Cleaning the Data

```{r pressure, echo=FALSE}
astronauts <- astronauts %>%
  mutate(occupation = tolower(occupation))

occupation_counts <- astronauts %>%
  count(occupation, sort = TRUE)


inorbit_counts <- astronauts %>%
  count(in_orbit, sort = TRUE)

unique_inorbit <- unique(astronauts$in_orbit)
# Convert all in_orbit values to the same title format and remove numbers behind
astronauts <- astronauts %>%
  mutate(in_orbit = gsub("[0-9]", "", in_orbit)) %>%
  mutate(in_orbit = gsub("-", " ", in_orbit))


# Remove extra spaces after "STS" in the in_orbit column
astronauts <- astronauts %>%
  mutate(in_orbit = gsub("STS\\s+", "STS ", in_orbit))


# Remove all non-alphanumeric characters except spaces
astronauts <- astronauts %>%
  mutate(in_orbit = gsub("[^[:alnum:]\\s]", "", in_orbit))

# Remove extra spaces
astronauts <- astronauts %>%
  mutate(in_orbit = gsub("\\s+", " ", in_orbit))



# Convert any value in the in_orbit column that includes "STS" to "STS"
astronauts <- astronauts %>%
  mutate(in_orbit = gsub(".*STS.*", "STS", in_orbit)) %>%
  mutate(in_orbit = gsub(".*Soyuz.*", "Soyuz", in_orbit)) %>%
  mutate(in_orbit = gsub(".*[Aa]pollo.*", "Apollo", in_orbit)) %>%
  mutate(in_orbit = gsub(".*[Gg]emini.*", "Gemini", in_orbit)) %>%
  mutate(in_orbit = gsub(".*Mir.*", "Mir", in_orbit)) %>%
  mutate(in_orbit = gsub(".*Saluyt*", "Salyut", in_orbit)) 


astronauts %>%
  count(in_orbit, sort = TRUE)
# Define a function to group in_orbit values
group_in_orbit <- function(in_orbit) {
  if (in_orbit %in% c("ISS", "Mir")) {
    return("Group 1")
  } else if (in_orbit %in% c("Salyut", "Skylab")) {
    return("Group 2")
  } else {
    return("Group 3")
  }
}

# Apply the function to create a new column for grouping
astronauts <- astronauts %>%
  mutate(in_orbit_group = sapply(in_orbit, group_in_orbit))

# Check the unique values in the new in_orbit_group column
unique(astronauts$in_orbit_group)
```

```{r}
# Filter out rows with 0 hours_mission
astronauts <- astronauts %>%
  filter(hours_mission != 0)

# Log-transform the hours_mission variable
log_transformed_astronauts <- astronauts %>%
  mutate(log_hours_mission = log(hours_mission))  # Adding 1 to avoid log(0)

```

```{r}
model_h <- stan_glm(
  log_hours_mission ~ 1 + in_orbit_group, data = log_transformed_astronauts,
  family = gaussian,
  chains = 4, iter = 5000*2, seed = 84735, refresh = 0)
```

```{r}
pp_check(model_h)
```

```{r}
tidy(model_h, conf.int = TRUE, conf.level = 0.8)
```

## Methodology
We are using hierarchical modeling to make predictions on the the mission hours based on multiple predictors, such as the orbit group and the occupation of each astronaut. 
The hierarchical model from the previous report (report 1) is not good enough so we split the data into 3 groups (that corresponds with different types of orbits) and we have a hierarchical model for them. The first two groups (group 1 and group 2) did not provide good information, and the model seemed more wrong. The last group, however, seems to be going well and this is because of the outliers. Therefore, going forward we will focus on group 3. 

## Bibliography

https://juliasilge.com/blog/astronaut-missions-bagging/

```{r}

log_transformed_astronauts$in_orbit <- as.factor(log_transformed_astronauts$in_orbit)

# complete_pooled <- stan_glmer(
#  log_hours_mission ~ (1 | in_orbit), data = log_transformed_astronauts,
#   family = gaussian,
#   chains = 4, iter = 1000*2, seed = 84735, refresh = 0)

saveRDS(complete_pooled, "models/hierachical.rds")

complete_pooled <- readRDS("models/hierachical.rds")


astronaut_means <- log_transformed_astronauts %>%
  group_by(in_orbit) %>%
  summarize(count = n(), mission = mean(log_hours_mission))
set.seed(84735)

predictions_complete <-  posterior_predict(
  complete_pooled, newdata = astronaut_means)

ppc_intervals(astronaut_means$mission, yrep = predictions_complete,
              prob_outer = 0.80) +
  ggplot2::scale_x_continuous(
    labels = astronaut_means$in_orbit,
    breaks = 1:nrow(astronaut_means)) +
    xaxis_text(angle = 90, hjust = 1)

pp_check(complete_pooled)

set.seed(84735)
prediction_summary(model = complete_pooled, data =log_transformed_astronauts)
```
over estimate the observations (between 5 - 7.5) and under estimate the peaks
```{r}
set.seed(84735)
predictions_complete_2 <- posterior_predict(
  complete_pooled,
  newdata = data.frame(in_orbit = c("Random", "STS")))

# Plot the posterior predictive models
mcmc_areas(predictions_complete_2, prob = 0.8) +
  xlab("hours") +
  ggplot2::scale_y_discrete(labels = c("Random", "STS"))

```
```{r}
predict_cv <- prediction_summary_cv(model = complete_pooled, group = "in_orbit", data = log_transformed_astronauts, k = 2)

saveRDS(complete_pooled, "models/hierachical.rds")
log_transformed_astronauts %>% head()
```

